Parameter Inference\\with Bifurcation Diagrams

### 2022-03-20 docs/appendix.tex

\clearpage
\section{Bifurcation Measure Properties}
\label{appendix:conditions}
Consider a vector $v(s)\in\Reals^N$ parameterised by $s\in\Reals$ that is tangent to an equilibrium manifold defined by $\rates(u)=0$. The conditions for a non-degenerate static bifurcation at $s^*$ along such a tangent can be expressed in terms of an eigenvalue $\lambda(s)$ of the state-space Jacobian crossing zero with a finite slope. A bifurcation exists at $s^*$ if
\begin{align}
    \frac{\partial\rates}{\partial u}v(s)=\lambda(s) \,v(s)
    \quad\exists\lambda:\quad
    \left.\lambda(s)\right|_{s=s^*}=0
    \qquad
    \left.\frac{d\lambda}{ds}\right|_{s=s^*}\neq 0
    \label{eq:conditions}
\end{align}
These conditions are necessary and sufficient for a non-degenerate static local breakdown of stability. For now we do not consider dynamic bifurcations involving limit cycles or imaginary parts of eigenvalues and restrict $\lambda\in\Reals$. Cases where both $\left.\lambda(s)\right|_{s=s^*}=0$ and $\left.\frac{d\lambda}{ds}\right|_{s=s^*}=0$ require investigation into higher order derivatives $\frac{d^n\lambda}{ds^n}$. These are the cases we refer to as \emph{degenerate} and are not considered here.

Instead of considering conditions on each eigenvalue individually it is possible to use the determinant of the state-space Jacobian to detect whether the conditions \eqref{eq:conditions} are satisfied. The determinant can be expressed as the product of eigenvalues
\begin{equation}
    \Det=\prod_{n=1}^N\lambda_n(s)
    \label{eq:determinant}
\end{equation}
Applying the product rule when differentiating yields
\begin{align}
    \frac{d}{ds}\Det&=
    \sum_{n=1}^N\frac{d\lambda_n}{ds}\prod_{n'\neq n}\lambda_{n'}(s)\\
    &=\Det\sum_{n=1}^N\frac{d\lambda_n}{ds}\lambda_{n}(s)^{-1}
\end{align}
Substituting this expression into measure \eqref{eq:measure}
\begin{equation}
    \measure(s)=
    \left(1+\left|\sum_{n=1}^N\frac{d\lambda_n}{ds}\lambda_{n}(s)^{-1}\right|^{-1}\right)^{-1}
\end{equation}
Which implies the following
\begin{equation}
    \exists\lambda:\quad
    \begin{cases}
        \,\lambda(s)=0 \quad\frac{d\lambda}{ds}\neq 0\\
        \,\lambda(s)\neq0 \quad\frac{d\lambda}{ds}\rightarrow\pm\infty
    \end{cases}
    \implies
    \measure(s)=1
\end{equation}
If there exists an eigenvalue that satisfies conditions \eqref{eq:conditions} then the measure is equal to one. The measure also approaches one in cases where the rate of change of an eigenvalue with respect to a manifold $s$ location  diverges while not crossing zero. This gives rise to finite gradients in the eigenvalue term in regimes far away from any bifurcation.

\section{Extension for Hopf Bifurcations}
\label{appendix:hopf}

@@

Suppose there exists a one dimensional space curve $\mathcal{C(\theta)}$ embedded in $z\in\Reals^{N+1}$ whose geometry changes depending on input parameters $\theta\in\Reals^M$. This curve could be open or closed and changes in $\theta$ could change the curve topology as well. Let the function $\gamma_{\theta}:\Reals\rightarrow\Reals^{N+1}$ be a parameterisation of the position vector along the curve within a fixed domain $s\in\mathcal{S}$. Note that the choice of parameterisation is arbitrary and our results should not depend on this choice. Furthermore, if we parametrise the curve $\mathcal{C}(\theta)$ with respect to a fixed domain $\mathcal{S}$ the dependence on $\theta$ is picked up by the parameterisation $\gamma_{\theta}(s)$. We can write a line integral of any scalar function $L_{\theta}:\Reals^{N+1}\rightarrow\Reals$ on the curve as

@@

The tangential component of the deformation field is not uniquely determined because there is no unique way of parameterising a surface. This is the subject of many computer graphics papers \cite{Jos2011OnSurface,Tao2016Near-IsometricTracking,Fujisawa2013CalculationInvariance}. We are however not interested in the continuous propagation of a mesh - as is the subject of those papers. In fact we are looking for a deformation field that is orthogonal to the tangent vector $\hat{\tangent}(z) \cdot\frac{d z}{d\theta} =0$ for the space curve, and therefore letting the tangential component of the deformation equal zero is a valid choice and we can it instead of the parameterised deformation

@@

\section{A More Complex Model}
\label{appendix:double-exclusive}
To demonstrate the wider reaching applicability of our method we optimise the \emph{double exclusive reporter} \cite{Grant2020InterpretationCircuit}, a synthetic \emph{E. Coli} designed to exhibit a cusp bifurcation with respect to two input signals. Let one of the input signals be the control condition $p$ while the other input is held constant at $1$. Figure \ref{fig:double-exclusive-optima} shows the cost function with respect to iteration for sample trajectories. We note that due to a

\begin{figure}[ht]
\centering
\setlength\unitlength{1cm}
{\phantomsubcaption\label{fig:double-exclusive-optima:parameters}}
{\phantomsubcaption\label{fig:double-exclusive-optima:models}}
\includegraphics[width=13cm]{double-exclusive-optima}
\begin{picture}(0,0) \put(-13,7){\subref{fig:double-exclusive-optima:parameters}} \end{picture}
\begin{picture}(0,0) \put(-7,7){\subref{fig:double-exclusive-optima:models}}
\end{picture}
\caption{Bifurcation inference for the \emph{double exclusive reporter}. \subref{fig:double-exclusive-optima:parameters}. Optimal parameter estimates $\theta^*$ for the targets $\targets=\{4,5\}$ reveal four clusters of qualitatively different regimes: mutual activation ($a_1 < 1$; cluster 1) and mutual inhibition ($a_1 > 1$; cluster 2). \subref{fig:double-exclusive-optima:models}. Example bifurcation diagrams indicate positively and negatively correlated dependencies between the two model states, as a function of the control condition.}
\label{fig:double-exclusive-optima}
\end{figure}

\section{Steady State Solvers}
\label{appendix:stedy-state-solvers}

>>>>

\section{Bifurcation Measure Properties}
\label{appendix:conditions}
Consider a vector $v(s)\in\Reals^N$ parametrised by $s\in\Reals$ that is tangent to an equilibrium manifold defined by $\rates(u)=0$. The conditions for a non-degenerate static bifurcation at $s^*$ along such a tangent can be expressed in terms of an eigenvalue $\lambda(s)$ of the state-space Jacobian crossing zero with a finite slope. A bifurcation exists at $s^*$ if
\begin{align}
    \frac{\partial\rates}{\partial u}v(s)=\lambda(s) \,v(s)
    \quad\exists\lambda:\quad
    \left.\lambda(s)\right|_{s=s^*}=0
    \qquad
    \left.\frac{d\lambda}{ds}\right|_{s=s^*}\neq 0
    \label{eq:conditions}
\end{align}
These conditions are necessary and sufficient for a non-degenerate static local breakdown of stability. For now we do not consider dynamic bifurcations involving limit cycles or imaginary parts of eigenvalues and restrict $\lambda\in\Reals$. Cases where both $\left.\lambda(s)\right|_{s=s^*}=0$ and $\left.\frac{d\lambda}{ds}\right|_{s=s^*}=0$ require investigation into higher order derivatives $\frac{d^n\lambda}{ds^n}$. These are the cases we refer to as \emph{degenerate} and are not considered here.

Instead of considering conditions on each eigenvalue individually it is possible to use the determinant of the state-space Jacobian to detect whether the conditions \eqref{eq:conditions} are satisfied. The determinant can be expressed as the product of eigenvalues
\begin{equation}
    \Det=\prod_{n=1}^N\lambda_n(s)
    \label{eq:determinant}
\end{equation}
Applying the product rule when differentiating yields
\begin{align}
    \frac{d}{ds}\Det&=
    \sum_{n=1}^N\frac{d\lambda_n}{ds}\prod_{n'\neq n}\lambda_{n'}(s)\\
    &=\Det\sum_{n=1}^N\frac{d\lambda_n}{ds}\lambda_{n}(s)^{-1}
\end{align}
Substituting this expression into measure \eqref{eq:measure}
\begin{equation}
    \measure(s)=
    \left(1+\left|\sum_{n=1}^N\frac{d\lambda_n}{ds}\lambda_{n}(s)^{-1}\right|^{-1}\right)^{-1}
\end{equation}
Which implies the following
\begin{equation}
    \exists\lambda:\quad
    \begin{cases}
        \,\lambda(s)=0 \quad\frac{d\lambda}{ds}\neq 0\\
        \,\lambda(s)\neq0 \quad\frac{d\lambda}{ds}\rightarrow\pm\infty
    \end{cases}
    \implies
    \measure(s)=1
    \label{eq:measure-conditions}
\end{equation}
If there exists an eigenvalue that satisfies conditions \eqref{eq:conditions} then the measure is equal to one. The measure also approaches one in cases where the rate of change of an eigenvalue with respect to a manifold $s$ location  diverges while not crossing zero. This gives rise to finite gradients in the eigenvalue term in regimes far away from any bifurcation.

@@

Suppose there exists a one dimensional space curve $\mathcal{C(\theta)}$ embedded in $z\in\Reals^{N+1}$ whose geometry changes depending on input parameters $\theta\in\Reals^M$. This curve could be open or closed and changes in $\theta$ could change the curve topology as well. Let the function $\gamma_{\theta}:\Reals\rightarrow\Reals^{N+1}$ be a parametrisation of the position vector along the curve within a fixed domain $s\in\mathcal{S}$. Note that the choice of parametrisation is arbitrary and our results should not depend on this choice. Furthermore, if we parametrise the curve $\mathcal{C}(\theta)$ with respect to a fixed domain $\mathcal{S}$ the dependence on $\theta$ is picked up by the parametrisation $\gamma_{\theta}(s)$. We can write a line integral of any scalar function $L_{\theta}:\Reals^{N+1}\rightarrow\Reals$ on the curve as

@@

The tangential component of the deformation field is not uniquely determined because there is no unique way of parametrising a surface. This is the subject of many computer graphics papers \cite{Jos2011OnSurface,Tao2016Near-IsometricTracking,Fujisawa2013CalculationInvariance}. We are however not interested in the continuous propagation of a mesh - as is the subject of those papers. In fact we are looking for a deformation field that is orthogonal to the tangent vector $\hat{\tangent}(z) \cdot\frac{d z}{d\theta} =0$ for the space curve, and therefore letting the tangential component of the deformation equal zero is a valid choice and we can it instead of the parametrised deformation

@@

\clearpage

\section{Application of Bifurcation Inference to a Complex Model}
\label{appendix:double-exclusive}

To demonstrate the wider reaching applicability of our method we optimise the \emph{double exclusive reporter} \cite{Grant2020InterpretationCircuit}, a synthetic gene circuit in \emph{E. coli} that was designed to exhibit a cusp bifurcation. The circuit behaviour is observed by measuring a fluorescent protein whose expression is controlled by transcription factors (regulatory proteins) LacI $(L)$ and TetR $(T)$, whose expression is in turn controlled by externally controllable \emph{input} signals $c_{6}$ and $c_{12}$. To apply the method, we consider one of the input signals be the control condition $c_{6}=p$, with the other packed together with the remaining 20 parameters into vector $\theta$. Once the optima $\theta^*$ have been obtained, we perform dimensionality reduction using \texttt{GigaSOM.jl} \cite{Kratochvil2020GigaSOM.jl:Datasets} so that the results can be visualised in a two dimensional embedding (Figure \ref{fig:double-exclusive-optima:parameters}).

The embedding reveals four optimal parameter regions. We find that, as with the two-state model in the main text \eqref{eq:two-state}, there are two qualitatively distinct regimes: mutual activation (region 1) and inhibition (regions 2-4). The mutual inhibition region can be further subdivided into three regions that are geometrically equivalent, but kinetically distinct: region 3 has swapped kinetic roles for regulatory proteins LacI and TetR compared to region 2, and region 4 has additional damped oscillations in the dynamics across the whole range of \emph{input} $c_{6}$ (Figure \ref{fig:double-exclusive-optima:models}). The two dimensional embedding of sampled optima $\theta^*$ enables navigation the space of qualitative behaviours of the \emph{double exclusive reporter} and organisation in terms of geometric and kinetic equivalence.

\begin{figure}[ht]
\centering
\setlength\unitlength{1cm}
{\phantomsubcaption\label{fig:double-exclusive-optima:parameters}}
{\phantomsubcaption\label{fig:double-exclusive-optima:models}}
\includegraphics[width=13cm]{double-exclusive-optima}
\begin{picture}(0,0) \put(-13,7){\subref{fig:double-exclusive-optima:parameters}} \end{picture}
\begin{picture}(0,0) \put(-7,7){\subref{fig:double-exclusive-optima:models}}
\end{picture}
\caption{Bifurcation inference for the \emph{double exclusive reporter}. \subref{fig:double-exclusive-optima:parameters}. Optimal parameter estimates $\theta^*$ for the targets $\targets=\{1,2\}$ (indicated by yellow lines in panel B) reveal four regions  with two geometrically different regimes: mutual activation (region 1) and mutual inhibition (regions 2-4). \subref{fig:double-exclusive-optima:models}. Example bifurcation diagrams indicate that region 2 has swapped kinetics between $L$ and $T$ to region 3. Region 4 has models with non-zero imaginary parts to eigenvalues indicating damped oscillations (shown in light green).}
\label{fig:double-exclusive-optima}
\end{figure}

These results were obtained with a modification of the bifurcation measure \eqref{eq:measure} to improve convergence rates. In parameter regimes where bifurcations are not present, according to conditions \eqref{eq:measure-conditions}, maximising the measure $\measure(s)$ can lead to a divergence in directional derivative $\frac{d\lambda}{ds}\rightarrow\pm\infty$ rather than a creation of a bifurcation. To discourage this from happening we can flatten out the gradients in that regime by applying the $\tanh$ non-linearity to the determinant. This leads to
\begin{align}
    \measure(s):=
    \left(1+\left|\frac{\tanh\Det}{\frac{d}{ds}\tanh\Det}\right|\right)^{-1}
    \label{eq:measure-tanh}
\end{align}

\clearpage
\section{Extension for Hopf Bifurcations}
\label{appendix:hopf}
